---
title: "Strawberry Analysis"
author: "Ruoxi Li"
format: html
editor: visual
---

## Data Set

The dataset contains strawberry farming data with details about conventional and organic cultivation.

## Objective

1.  Pick three chemical treatments used for conventional strawberries in both states and contrast their use. Try to find chemicals with divergent use patterns between the states. Produce, tables, plots, and descriptions of the chemicals, how they are used, and how their use differs between California and Florida.

2.  Compare the production and sales of organic and conventional strawberries and strawberries sold for processing. Show differences in price and volume between California and Florida. How do price, cost, and volume relationships change over the years?

Install packages

```{r}
# install required packages
#install.packages("knitr")
#install.packages("kableExtra")
#install.packages("tidyverse")
#install.packages("stringr")
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("scales")

# loading required packages
library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
library(ggplot2)
library(dplyr)
library(scales)
library(tidyr)
```

## Question 1:

## Methodology

This analysis used USDA survey data from 2021-2023 to compare the application of chemical treatments in conventional strawberry production between California and Florida. By examining the `Domain Category` field in the dataset, we identified chemicals with the most divergent usage patterns between the two states. The analysis focuses on three chemicals that show particularly distinct usage patterns:

1.  **Captan** (fungicide)

2.  **Spinetoram** (insecticide)

3.  **Acetamiprid** (insecticide)

```{r}
# Chemical Treatment Analysis for Strawberries
# Comparing usage patterns between California and Florida

# Read the data file
strawberry <- read_csv("strawb_mar6.csv", 
                       col_names = TRUE,
                       show_col_types = FALSE)

# Filter for chemical data in California and Florida
chem_data <- strawberry %>%
  filter(
    State %in% c("CALIFORNIA", "FLORIDA"),
    grepl("CHEMICAL", Domain)
  ) %>%
  # Extract chemical names and codes from Domain Category
  mutate(
    ChemicalName = case_when(
      grepl("CAPTAN = 81301", `Domain Category`) ~ "Captan",
      grepl("SPINETORAM = 110007", `Domain Category`) ~ "Spinetoram",
      grepl("ACETAMIPRID = 99050", `Domain Category`) ~ "Acetamiprid",
      TRUE ~ NA_character_
    )
  ) %>%
  # Keep only our target chemicals
  filter(!is.na(ChemicalName)) %>%
  # Clean up Value column - remove commas and convert to numeric
  mutate(
    Value = as.numeric(gsub(",", "", Value)),
    # Create a clean chemical type variable
    ChemicalType = case_when(
      grepl("FUNGICIDE", Domain) ~ "Fungicide",
      grepl("INSECTICIDE", Domain) ~ "Insecticide",
      grepl("HERBICIDE", Domain) ~ "Herbicide",
      grepl("OTHER", Domain) ~ "Other Chemical",
      TRUE ~ "Unknown"
    )
  )

# Function to check if data contains (D) values, which indicate disclosure concerns
has_disclosure_issues <- function(data) {
  any(grepl("\\(D\\)", data$Value))
}

# Check if we have disclosure concerns
if(has_disclosure_issues(chem_data)) {
  cat("Note: Some values contain (D), indicating they are withheld for disclosure concerns.")
}

# Summarize chemical usage by state, year, and chemical
chem_summary <- chem_data %>%
  group_by(State, Year, ChemicalName, ChemicalType) %>%
  summarize(
    AvgValue = mean(Value, na.rm = TRUE),
    .groups = "drop"
  )

# Create tables showing usage by state and year
state_chem_table <- chem_summary %>%
  pivot_wider(
    names_from = Year,
    values_from = AvgValue,
    names_prefix = "Year_"
  ) %>%
  arrange(ChemicalName, State)

# Calculate overall averages by chemical and state
state_chem_avg <- chem_summary %>%
  group_by(State, ChemicalName, ChemicalType) %>%
  summarize(
    AvgUsage = mean(AvgValue, na.rm = TRUE),
    .groups = "drop"
  )

# Create a wide format table for comparison
state_comparison <- state_chem_avg %>%
  pivot_wider(
    names_from = State,
    values_from = AvgUsage
  ) %>%
  mutate(
    Difference = CALIFORNIA - FLORIDA,
    PercentDiff = (Difference / ((CALIFORNIA + FLORIDA)/2)) * 100
  )

# Create plot for chemical comparison between states
ggplot(state_chem_avg, aes(x = ChemicalName, y = AvgUsage, fill = State)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Average Chemical Usage in Strawberry Production",
    subtitle = "Comparison between California and Florida",
    x = "Chemical",
    y = "Usage Value",
    fill = "State",
    caption = "Source: USDA Strawberry Production Data"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
ggsave("chemical_comparison.png", width = 8, height = 6)

# Create plot for time trends by chemical and state
ggplot(chem_summary, aes(x = Year, y = AvgValue, color = State, group = State)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  facet_wrap(~ChemicalName, scales = "free_y") +
  labs(
    title = "Chemical Usage in Strawberry Production Over Time",
    subtitle = "Trends in California vs. Florida (2021-2023)",
    x = "Year",
    y = "Usage Value",
    color = "State",
    caption = "Source: USDA Strawberry Production Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
ggsave("chemical_trends.png", width = 10, height = 8)

# Get additional information about each chemical from the data
chemical_info <- tibble(
  ChemicalName = c("Captan", "Spinetoram", "Acetamiprid"),
  PrimaryUse = c("Fungicide", "Insecticide", "Insecticide"),
  Description = c(
    "A protective fungicide used to control diseases like anthracnose, Botrytis, and other fungal pathogens in strawberries.",
    "A spinosyn insecticide that targets pests like thrips, leaf miners, and caterpillars through contact and ingestion.",
    "A neonicotinoid insecticide that controls aphids, whiteflies, and other sucking pests in strawberry production."
  ),
  UsagePattern = c(
    "Higher usage in California, likely due to favorable conditions for fungal pathogens in the cool, coastal growing regions.",
    "Significantly higher usage in Florida, possibly reflecting different pest pressures and management approaches.",
    "Much higher usage in Florida, suggesting different insect pressure or resistance management strategies."
  )
)

# Print tables
cat("\n--- Chemical Usage Comparison between California and Florida ---\n")
print(state_comparison)

cat("\n--- Chemical Usage by State and Year ---\n")
print(state_chem_table)

cat("\n--- Chemical Information ---\n")
print(chemical_info)

# Additional analysis: Look for correlations between chemical use and environmental factors
# Note: This would require merging with weather/climate data not present in the dataset

# Create a comprehensive report table
report_table <- chemical_info %>%
  left_join(state_comparison, by = "ChemicalName") %>%
  select(ChemicalName, PrimaryUse, CALIFORNIA, FLORIDA, Difference, PercentDiff, Description, UsagePattern) %>%
  rename(
    Chemical = ChemicalName,
    `California Usage` = CALIFORNIA,
    `Florida Usage` = FLORIDA,
    `Absolute Difference` = Difference,
    `Percent Difference` = PercentDiff,
    `Primary Purpose` = PrimaryUse
  )

# Format the report table nicely
kable(report_table, format = "html", caption = "Chemical Treatments in Strawberry Production: California vs. Florida") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

# Output this table for the report
write.csv(report_table, "chemical_treatment_comparison.csv", row.names = FALSE)

# Create a chart showing relative usage proportions for each chemical by state
prop_data <- state_chem_avg %>%
  group_by(State) %>%
  mutate(Proportion = AvgUsage / sum(AvgUsage) * 100)

ggplot(prop_data, aes(x = State, y = Proportion, fill = ChemicalName)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Proportion of Chemical Usage by State",
    subtitle = "Relative contribution of each chemical to total usage",
    x = "State",
    y = "Percentage of Total Usage",
    fill = "Chemical",
    caption = "Source: USDA Strawberry Production Data"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
ggsave("chemical_proportions.png", width = 8, height = 6)

# Summary statistics and analysis notes
cat("\n--- Key Findings ---\n")
cat("1. Captan (fungicide) is used approximately", round(state_comparison$PercentDiff[state_comparison$ChemicalName == "Captan"]), 
    "% more in California than in Florida.\n")
cat("2. Spinetoram (insecticide) shows dramatically higher usage in Florida, with approximately", 
    abs(round(state_comparison$PercentDiff[state_comparison$ChemicalName == "Spinetoram"])), 
    "% more use compared to California.\n")
cat("3. Acetamiprid (insecticide) is also significantly more used in Florida, with about", 
    abs(round(state_comparison$PercentDiff[state_comparison$ChemicalName == "Acetamiprid"])), 
    "% more use than in California.\n")
cat("4. These differences likely reflect regional variations in pest pressure, climate conditions,\n",
    "   and different integrated pest management strategies between the two major strawberry-producing states.\n")
```

## Chemical Profiles and Usage Patterns

### Captan (Fungicide)

**Chemical Properties and Purpose:** Captan is a broad-spectrum, protective fungicide that has been used in agriculture since the 1950s. It works by inhibiting spore germination and preventing fungal growth on plant surfaces. In strawberry production, Captan is primarily used to control anthracnose, Botrytis gray mold, leaf spot, and various other fungal diseases that can significantly reduce yield and fruit quality.

**Usage Pattern Comparison:**

-   **California:** Higher usage (approximately 104 units)

-   **Florida:** Lower usage (approximately 49 units)

-   **Difference:** 55 units more in California (approximately 72% higher)

**Analysis:** The substantially higher use of Captan in California reflects several regional factors:

-   California's coastal growing regions often experience cooler, foggier conditions that create ideal environments for fungal diseases, particularly Botrytis gray mold.

-   The longer growing season in California may necessitate more frequent fungicide applications throughout the year.

-   California's production system often employs raised beds with plastic mulch, which can create microclimates favorable to certain fungal pathogens.

-   Resistance management strategies in California may rely more heavily on Captan as a multi-site fungicide that faces less resistance pressure than newer single-site fungicides.

### Spinetoram (Insecticide)

**Chemical Properties and Purpose:** Spinetoram is a second-generation spinosyn insecticide derived from the fermentation of Saccharopolyspora spinosa, a naturally occurring soil bacterium. It affects the insect's nervous system through a unique mode of action, causing paralysis and death. In strawberry production, Spinetoram targets pests such as thrips, leaf miners, caterpillars, and other lepidopteran pests that damage foliage and fruit.

**Usage Pattern Comparison:**

-   **California:** Lower usage (approximately 13 units)

-   **Florida:** Higher usage (approximately 152 units)

-   **Difference:** 139 units more in Florida (approximately 169% higher)

**Analysis:** The dramatically higher use of Spinetoram in Florida indicates significant differences in pest management approaches:

-   Florida's warmer climate supports year-round pest activity, particularly thrips and lepidopteran species that Spinetoram targets effectively.

-   Florida's strawberry production faces greater pressure from specific pests like chilli thrips (*Scirtothrips dorsalis*) and western flower thrips (*Frankliniella occidentalis*), which have developed resistance to other insecticides.

-   The winter growing season in Florida corresponds with peak periods for certain insect pests that require targeted control.

-   Regulatory differences or resistance management programs in Florida may favor Spinetoram over alternative insecticides.

### Acetamiprid (Insecticide)

**Chemical Properties and Purpose:** Acetamiprid is a neonicotinoid insecticide that acts as a nicotinic acetylcholine receptor agonist, disrupting the insect's nervous system. It is systemic, meaning plants absorb and distribute it throughout their tissues. In strawberry production, Acetamiprid primarily controls aphids, whiteflies, thrips, and other sucking pests that can vector viral diseases and reduce plant vigor.

**Usage Pattern Comparison:**

-   **California:** Lower usage (approximately 8 units)

-   **Florida:** Higher usage (approximately 104 units)

-   **Difference:** 96 units more in Florida (approximately 171% higher)

**Analysis:** The significantly higher use of Acetamiprid in Florida strawberry production suggests:

-   Florida faces greater pressure from aphids and whiteflies that thrive in warmer climates.

-   Vector management may be a higher priority in Florida due to the presence of specific viral diseases transmitted by aphids.

-   Regulatory restrictions or resistance management strategies in California may limit neonicotinoid use more stringently.

-   Florida growers may use Acetamiprid as part of a rotation strategy with other insecticides to manage resistance development.

## Comparative Analysis and Implications

The stark differences in chemical usage between California and Florida demonstrate how regional environmental factors, pest pressures, and management strategies influence strawberry production practices:

1.  **Climate Impact:** California's coastal, Mediterranean climate versus Florida's subtropical conditions create fundamentally different disease and pest pressures, driving divergent chemical management strategies.

2.  **Growing Season Influence:** Florida's winter production cycle versus California's extended growing season necessitates different timing and frequency of chemical applications.

3.  **Regulatory Framework:** Variations in state regulations regarding chemical use likely influence treatment selections and application rates.

4.  **Resistance Management:** Different regional histories of chemical use have likely led to varying pest resistance patterns, necessitating region-specific management strategies.

5.  **Cost-Benefit Considerations:** The economic thresholds for chemical intervention may differ between the two production regions based on market timing, labor costs, and expected returns.

## Question 2:

Read file

```{r}
# The following code was adapted from https://github.com/MA615-415-Spring-2025/class-15---Mar-20.git
# Label for the code chunk
#| label: read data - glimpse 

# Read the CSV file "strawb_mar6.csv" into a data frame called 'strawberry'
# - 'col_names = TRUE' tells read_csv to treat the first row as column names
# - 'show_col_types = FALSE' suppresses printing of column type information
strawberry <- read_csv("strawb_mar6.csv", 
                       col_names = TRUE,
                       show_col_types = FALSE)

# Source the file "my_functions.R" to load custom functions defined in that script
source("my_functions.R")
```

```{r}
# The following code was adapted from https://github.com/MA615-415-Spring-2025/class-15---Mar-20.git
# remove columns from a data frame that contain only a single unique value
strawb <- strawberry |> drop_one_value_col()
```

Exploring Data

```{r}
#| label: explore strawb data
# The following code was adapted from https://github.com/MA615-415-Spring-2025/class-15---Mar-20.git
# assume data is a tibble
# n_show is the number of rows to show

show_unique <- function(data, nrows=10 ){
  # make a tibble items to hold the data to show
  # browser()
  a <- nrows * dim(data)[2]  # number of cells in items
  items <- rep(" ", a) # items will coerce everything to char
  dim(items) <- c(nrows ,dim(data)[2]) # shape items
  items <- as_tibble(items)
  colnames(items) <- colnames(data)
  # browser()
  for(i in 1:dim(data)[2]){

    col_items <- unique(data[,i])
    # row_ex is the number of rows needed 
    # to make the column length conformable with items
    row_ex <- nrows - dim(col_items)[1] 
    if(row_ex >= 0){
      ex_rows <- tibble(rep(" ",row_ex))
      colnames(ex_rows) <- colnames(col_items)
      col_add <- rbind2(col_items, ex_rows)
      
    } else if(row_ex < 0){
      col_add <- col_items[1:10,]
      
    }

    items[,i] <- col_add

  }
  
  return(items)
}

#test <- show_unique(strawb, 10)
```

```{r}
# The following code was adapted from https://github.com/MA615-415-Spring-2025/class-15---Mar-20.git
#|label: split strawb into census and survey pieces

strw_census <- strawb |> filter(Program == "CENSUS")

strw_survey <- strawb |> filter(Program == "SURVEY")

nrow(strawb) == (nrow(strw_census) + nrow(strw_survey))
```

```{r}
# The following code was adapted from https://github.com/MA615-415-Spring-2025/class-15---Mar-20.git
#| label: examine Census and survey tibbles
#| warning: false

# Remove columns that contain only a single unique value
s_census <- strw_census |> drop_one_value_col(prt_val = TRUE)
s_survey <- strw_survey |> drop_one_value_col(prt_val = TRUE)

# Preview up to 10 unique values per column in each data set
unique_cen <- s_census |> show_unique(nrows = 10)
unique_sur <- s_survey |> show_unique(nrows = 10)

# Drop redundant or unnecessary columns from Census and Survey data
strw_census <- s_census |> select(-`State ANSI`)

# Drop 'State ANSI', 'Week Ending', and 'Period' for clarity and relevance
strw_survey <- s_survey |> select(-`State ANSI`, -`Week Ending`, -Period)

# Clean up workspace by removing intermediate and temporary objects
rm(s_census, s_survey, strawberry, strawb, items)
```

## Strawberry production location comparison

```{r}
# Combine Census and Survey data
all_strw <- bind_rows(strw_census, strw_survey)

# Count observations by State
state_all1 <- all_strw |> 
  group_by(State) |> 
  count() |> 
  arrange(desc(n))  # Sort by count

# Improved bar plot
ggplot(state_all1, aes(x = reorder(State, -n), y = n)) +
  geom_bar(stat = "identity", fill = "tomato", color = "black") +
  labs(
    title = "Strawberry Production by State",
    x = "State",
    y = "Number of Records"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

Based on the bar chart showing the number of records per state in the USDA strawberry dataset:

California is by far the leading state in strawberry production, with the highest number of records in the dataset. Florida comes in second, with a substantial number of entries as well. Other states with recorded strawberry activity include: Connecticut, Massachusetts, Maine, New Hampshire, Rhode Island, Vermont, New York.

```{r}
# Show unique values in the 'Domain' column from both datasets
unique(strw_census$Domain)
unique(strw_survey$Domain)
```

```{r}
# Count number of records per Domain in Census
strw_census |> 
  count(Domain, sort = TRUE)

# Count number of records per Domain in Survey
strw_survey |> 
  count(Domain, sort = TRUE)
```

```{r, fig.width = 10, fig.height = 6}
# Clean the Value column: remove commas and ensure numeric conversion only where possible
strw_census <- strw_census |> 
  mutate(Value = gsub(",", "", Value),           # Remove commas
         Value = as.numeric(Value))              # Convert to numeric

# Summarize by total strawberry production value by sales tier
farm_sales_value <- strw_census |>
  filter(Domain == "FARM SALES") |>
  group_by(`Domain Category`) |>
  summarise(total_value = sum(Value, na.rm = TRUE)) |>
  arrange(desc(total_value))

ggplot(farm_sales_value, aes(x = reorder(`Domain Category`, total_value), y = total_value)) +
  geom_bar(stat = "identity", fill = "darkorange", color = "black") +
  labs(title = "Strawberry Production Value by Farm Sales Tier",
       x = "Farm Sales Tier",
       y = "Total Production Value ($)") +
  scale_y_continuous(labels = label_comma()) +  # <- formats axis nicely
  coord_flip() +
  theme_minimal(base_size = 14)
```

According to USDA Census data, strawberries are primarily grown by large-scale commercial farms. The \$1,000,000+ sales tier alone accounts for the overwhelming majority of the total production value, highlighting that strawberry farming in the U.S. is dominated by high-output, industrial-scale operations.

While there are many small farms represented in the data, their economic impact is minimal by comparison. Mid-sized farms — those making between \$250,000 and \$999,999 — also contribute meaningfully, but their combined value still pales next to the top-tier producers.

```{r,fig.width = 10, fig.height = 6}
# Ensure numeric values
strw_census <- strw_census |>
  mutate(Value = as.numeric(gsub(",", "", Value)))

# Filter for certified organic sales
organic_sales_usd <- strw_census |> 
  filter(
    grepl("SALES", Category, ignore.case = TRUE),
    grepl("ORGANIC", Category, ignore.case = TRUE),
    Domain == "ORGANIC STATUS",
    `Domain Category` == "ORGANIC STATUS: (NOP USDA CERTIFIED)",
    !is.na(Value)
  )

# Summarize by state
organic_sales_usd_by_state <- organic_sales_usd |> 
  group_by(State) |> 
  summarise(total_sales_usd = sum(Value, na.rm = TRUE)) |> 
  arrange(desc(total_sales_usd))

# Prepare the percentage + label text
organic_sales_usd_by_state <- organic_sales_usd_by_state |> 
  mutate(
    pct = total_sales_usd / sum(total_sales_usd),
    label = paste0(State, ": ", scales::percent(pct, accuracy = 0.1))
  )

# Recalculate % and group small states
threshold <- 0.05

organic_sales_usd_by_state <- organic_sales_usd_by_state |>
  mutate(pct = total_sales_usd / sum(total_sales_usd)) |>
  mutate(State_grouped = ifelse(pct < threshold, "Other", State)) |>
  group_by(State_grouped) |>
  summarise(total_sales_usd = sum(total_sales_usd)) |>
  ungroup() |>
  mutate(
    pct = total_sales_usd / sum(total_sales_usd),
    label = paste0(State_grouped, ": ", percent(pct, accuracy = 0.1))
  )


# Add conditional label (only show if > 3%)
organic_sales_usd_by_state <- organic_sales_usd_by_state |> 
  mutate(
    show_label = ifelse(pct > 0.03, label, "")
  )
# Plot donut chart with combined categories
ggplot(organic_sales_usd_by_state, aes(x = "", y = total_sales_usd, fill = State_grouped)) +
  geom_bar(width = 0.6, stat = "identity", color = "black") +
  coord_polar("y", start = 0) +
  geom_text(
    aes(label = show_label), 
    position = position_stack(vjust = 0.5),
    size = 4,
    color = "white",
    fontface = "bold"
  ) +
  labs(
    title = "Certified Organic Strawberry Sales by State",
    fill = "State"
  ) +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

Certified organic strawberries are overwhelmingly produced by large-scale operations in California, with Florida contributing a smaller but notable share. Other states participate in organic production, but their total sales are minimal. This implies that organic strawberry farming is both geographically and economically concentrated, likely driven by economies of scale and climate advantages in California.

```{r}
# Clean data and create a new column "Measurement" that identifies if the row is for Acres or Yield
straw_data_clean <- all_strw %>%
  mutate(
    # Replace en-dash with hyphen and remove extra spaces
    Fruit = str_replace_all(Fruit, "–", "-"),
    Fruit = str_squish(Fruit),
    # Convert Value to numeric (non-numeric entries become NA)
    Value = as.numeric(gsub(",", "", Value)),
    # Create a new column based on keywords in Fruit
    Measurement = case_when(
      str_detect(Fruit, "ACRES HARVESTED") ~ "Acres",
      str_detect(Fruit, "YIELD") ~ "Yield",
      TRUE ~ NA_character_
    )
  )

# Filter for California and Florida and only rows that have a valid Measurement
volume_data <- straw_data_clean %>%
  filter(
    State %in% c("CALIFORNIA", "FLORIDA"),
    !is.na(Measurement)
  ) %>%
  select(Year, State, Measurement, Value)

# Pivot so that we get separate columns for Acres and Yield
volume_data <- volume_data %>%
  pivot_wider(names_from = Measurement, values_from = Value)

# Calculate production volume in CWT (assume Volume_CWT = Acres * Yield)
volume_data <- volume_data %>%
  mutate(Production_CWT = Acres * Yield)

# Check the resulting data frame
print(volume_data)

# Plot production volume over the years for California vs Florida
ggplot(volume_data, aes(x = Year, y = Production_CWT, color = State, group = State)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Strawberry Production Volume (CWT) by Year: CA vs FL",
    x = "Year",
    y = "Production Volume (CWT)"
  ) +
  theme_minimal(base_size = 14)
```

California consistently outproduces Florida, with \~5-8 times the total CWT. Over 2020–2023, California’s production hovers between 23–25 million CWT, while Florida ranges around 3–4 million CWT. Both states show relatively stable or slightly increasing volumes across these years.

```{r}
ggplot(volume_data, aes(x = Year, y = Yield, color = State, group = State)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Strawberry Yield (CWT / Acre) by Year: CA vs FL",
    x = "Year",
    y = "Yield (CWT / Acre)"
  ) +
  theme_minimal(base_size = 14)
```

California achieves a much higher yield (around 600–700 CWT/acre) than Florida (about 200–220 CWT/acre). California’s yield slightly declines over time (e.g., \~660 CWT/acre in 2020 down to \~620 by 2023), while Florida’s yield stays lower but is relatively stable or slightly up.

California’s climate, infrastructure, and farming practices give it a yield advantage. Florida’s lower yields may reflect soil/climate constraints or different farming methods.

```{r}
ggplot(volume_data, aes(x = Year, y = Acres, color = State, group = State)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Strawberry Acres Harvested by Year: CA vs FL",
    x = "Year",
    y = "Acres"
  ) +
  theme_minimal(base_size = 14)

```

California has around 38k–43k acres over 2020–2023, while Florida ranges from 6k–10k acres. Both states expand harvested acreage in the middle years, with California peaking around 43k in 2022 and Florida nearing 10k acres in 2023.

California’s larger land base dedicated to strawberries (combined with higher yields) explains its dominant production volume. Florida is increasing acreage over time but remains well below California.

```{r}
price_data <- straw_data_clean %>%
  filter(
    State %in% c("CALIFORNIA", "FLORIDA"),
    Fruit == "STRAWBERRIES - PRICE RECEIVED"
  )

price_summary <- price_data %>%
  group_by(Year, State) %>%
  summarise(
    AvgPrice = mean(Value, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(price_summary, aes(x = Year, y = AvgPrice, color = State, group = State)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Strawberry Price ($/CWT) by Year: CA vs FL",
    x = "Year",
    y = "Price ($ per CWT)"
  ) +
  theme_minimal(base_size = 14)
```

Florida commands a higher price per CWT than California in most years. Florida’s price ranges from \$140–\$170 (with a peak around 2022), then dips to \~\$120 in 2023. California’s price is lower, starting around \$90 in 2020, peaking near \$110–\$115, then settling to \~\$95 in 2023.

Florida’s smaller supply and potentially niche or local markets may keep prices higher. California’s large-scale production can drive prices down. Over time, price volatility is noticeable: Florida spikes in 2022, then drops; California sees a smaller range of fluctuation.

```{r}
# Focus on 2021 data and calculate organic price from sales values
organic_sales_dollars <- all_strw %>%
  filter(
    Year == 2021,
    State %in% c("CALIFORNIA", "FLORIDA"),
    Domain == "ORGANIC STATUS",
    `Domain Category` == "ORGANIC STATUS: (NOP USDA CERTIFIED)",
    grepl("SALES", Category),
    grepl("$", Item)
  ) %>%
  mutate(Value = as.numeric(gsub(",", "", Value))) %>%
  group_by(State) %>%
  summarise(OrgSalesDollars = sum(Value, na.rm = TRUE))

organic_sales_cwt <- all_strw %>%
  filter(
    Year == 2021,
    State %in% c("CALIFORNIA", "FLORIDA"),
    Domain == "ORGANIC STATUS",
    `Domain Category` == "ORGANIC STATUS: (NOP USDA CERTIFIED)",
    grepl("SALES", Category),
    grepl("MEASURED IN CWT", Item)
  ) %>%
  mutate(Value = as.numeric(gsub(",", "", Value))) %>%
  group_by(State) %>%
  summarise(OrgSalesCWT = sum(Value, na.rm = TRUE))

# Calculate organic price per CWT
organic_price_2021 <- organic_sales_dollars %>%
  inner_join(organic_sales_cwt, by = "State") %>%
  mutate(OrgPrice = OrgSalesDollars / OrgSalesCWT)

# Use the existing price_data table for conventional prices
conventional_price_2021 <- price_data %>%
  filter(
    Year == 2021,
    State %in% c("CALIFORNIA", "FLORIDA")
  ) %>%
  group_by(State) %>%
  summarise(ConvPrice = mean(Value, na.rm = TRUE))

# Calculate price premium using the organic_price_2021 obtained earlier
price_premium_2021 <- organic_price_2021 %>%
  inner_join(conventional_price_2021, by = "State") %>%
  mutate(
    PricePremium = (OrgPrice - ConvPrice) / ConvPrice * 100
  )

# Calculate price premium
price_premium_2021 <- organic_price_2021 %>%
  inner_join(conventional_price_2021, by = "State") %>%
  mutate(
    PricePremium = (OrgPrice - ConvPrice) / ConvPrice * 100
  )

# Create the visualization
ggplot(price_premium_2021, aes(x = State, y = PricePremium, fill = State)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("CALIFORNIA" = "#1f77b4", "FLORIDA" = "#ff7f0e")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Organic Price Premium for Strawberries (2021)",
    subtitle = "Percentage increase over conventional prices",
    x = "State",
    y = "Price Premium (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "none"
  )
```

Both states demonstrate substantial organic price premiums, indicating that consumers in both markets are willing to pay significantly more for organically grown strawberries. The premiums hover around 70-80% above conventional prices, which is notable even compared to typical organic premiums in other agricultural products.

California shows a premium of approximately 76%, while Florida exhibits a slightly lower but still substantial premium of about 71%. This indicates that despite geographical and market differences, organic strawberries maintain significant price advantages in both major production regions.

```{r,fig.width = 10, fig.height = 6}

census_organic_acres <- strw_census %>%
  mutate(
    # Convert Value to numeric
    Value = as.numeric(gsub(",", "", Value))
  ) %>%
  # Filter for year=2021, states CA/FL, and Category=ORGANIC - ACRES HARVESTED
  filter(
    Year == 2021,
    State %in% c("CALIFORNIA", "FLORIDA"),
    Category == " ORGANIC - ACRES HARVESTED"
  ) %>%
  # Summarize (if there are multiple rows, sum them up)
  group_by(Year, State) %>%
  summarise(Total_Acres_Census = sum(Value, na.rm = TRUE)) %>%
  ungroup()

survey_acres <- strw_survey %>%
  mutate(
    Value = as.numeric(gsub(",", "", Value))
  ) %>%
  # Filter for year=2021, states CA/FL, and Fruit=STRAWBERRIES - ACRES HARVESTED
  filter(
    Year == 2021,
    State %in% c("CALIFORNIA", "FLORIDA"),
    Fruit == "STRAWBERRIES - ACRES HARVESTED"
  ) %>%
  group_by(Year, State) %>%
  summarise(Total_Acres_Survey = sum(Value, na.rm = TRUE)) %>%
  ungroup()

acres_compare <- left_join(
  census_organic_acres,
  survey_acres,
  by = c("Year", "State")
)

acres_compare_long <- acres_compare %>%
  pivot_longer(
    cols = c("Total_Acres_Census", "Total_Acres_Survey"),
    names_to = "Source",
    values_to = "Acres"
  )

ggplot(acres_compare_long, aes(x = State, y = Acres, fill = Source)) +
  geom_col(position = "dodge") +
  labs(
    title = "Comparing Organic Acres (Orange) vs. All/Conventional Acres (Blue) in 2021",
    x = "State",
    y = "Acres"
  ) +
  theme_minimal(base_size = 14)
```

California: \~40k total acres (blue) vs. only a few thousand organic acres (orange).

Florida: \~10k total acres vs. a small fraction organic (\~1k).

Organic remains a small slice of total acreage in both states.

Even in the top strawberry-producing states, organic acreage is significantly smaller than conventional. California leads in absolute organic acres (simply because it has more farmland overall), but as a percentage, organic is still a minor share of total production.

From all the data and graph, we can reach a conclusion that the U.S. strawberry market reveals significant contrasts between California and Florida production systems. California dominates the industry with substantially greater acreage and volume, producing over 20 million CWT compared to Florida's 3-4 million CWT. This scale advantage likely contributes to California's relatively stable pricing range (\$90-\$115 per CWT) and potentially lower unit costs. Florida, despite its smaller production footprint, often commands higher prices for its strawberries, though with greater market volatility, as evidenced by price peaks in 2022 followed by declines in 2023. The organic segment remains modest in both states, constituting a small fraction of total production, though California maintains more organic acreage in absolute terms. While processing data isn't explicitly detailed, a portion of total production from both states typically goes toward processed products like jams and frozen strawberries, with processing prices generally differing from fresh market prices. Over time, California has maintained relatively consistent production volumes while Florida shows modest increases in acreage and production, though still at levels far below California's scale.

```{r}
#rm(list=ls())
citation()
```

## References

```         
R Core Team (2024). _R: A Language and Environment for Statistical Computing_. R Foundation for Statistical Computing, Vienna, Austria.
  <https://www.R-project.org/>.
```

### Dataset Source

Massachusetts Institute of Technology \[MIT\]. (2025). *Strawberry Analysis Dataset* \[Data file\]. Retrieved from <https://github.com/MA615-415-Spring-2025/class-15---Mar-20.git>

### AI Assistance

OpenAI. (2025). ChatGPT (GPT-o3 mini high) \[Large language model\]. Retrieved from [https://chat.openai.com](https://chat.openai.com/)

Anthropic. (2025). Claude 3.7 Sonnet \[Large language model\]. Retrieved from [https://claude.ai](https://claude.ai/)
